Many claims of group polarization are plausibly false due to inappropriate statistics

### 2022-11-21 GroupPolarizationStatmod/paper/ms.tex
added table with experiment conditions info and GP schematic figure

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{setspace}
\usepackage{longtable}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

\begin{abstract}
  \noindent
  Group polarization is the observed tendency of
  like-minded groups to arrive at a more extreme consensus opinion than
  the average opinion after discussing some topic of interest.
  Clearly it is important to understand such a phenomenon to understand broader
  problematic trends towards increased polariztion across society as a whole.
  Unfortunately, recent analyses have demonstrated that this measurement and
  analysis approach can result in false positive detections group polarization
  from failing to account for floor/ceiling effects induced when using ordinal scales to
  measure continuously-distributed latent psychological variables. 
  Here we demonstrate that many high profile detections
  of group polarization are plausibly false
  due to these floor/ceiling effects using generative modeling seeded by 
  reported group polarization values and parameters. After we identify which
  reported detections of group polarization are plausibly false detections, we further
  investigate whether the latent constant mean opinion and pre- and post-discussion
  opinion variances can fool the statistical tests used in each study to 
  generate false positives. Finally, we evaluate whether a more appropriate
  Bayesian cumulative normal statistical model can more accurately distinguish
  group polarization from simple consensus. Group polarization continues to 
  pose a problem for psychological theory with an ever-expanding set of potential theoretical
  explanations. A search for explanations is pointless, however, without sound
  evidence. We believe this work will provide a sturdier analytical foundation for 
  future group-polarization research.
\end{abstract}

@@

@@

>>>>

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{setspace}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{graphicx}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

\begin{abstract}
  \noindent
  Group polarization is the observed tendency of
  like-minded groups to arrive at a more extreme consensus opinion than
  the average opinion after discussing some topic of interest.
  Clearly it is important to understand such a phenomenon to understand broader
  problematic trends towards increased polariztion across society as a whole.
  Unfortunately, recent analyses have demonstrated that this measurement and
  analysis approach can result in false positive detections group polarization
  from failing to account for floor/ceiling effects induced when using ordinal scales to
  measure continuously-distributed latent psychological variables. 
  Here we demonstrate that many high profile detections
  of group polarization are plausibly false
  due to these floor/ceiling effects using generative modeling seeded by 
  reported group polarization values and parameters. After we identify which
  reported detections of group polarization are plausibly false detections, we further
  investigate whether the latent constant mean opinion and pre- and post-discussion
  opinion variances can fool the statistical tests used in each study to 
  generate false positives. Finally, we evaluate whether a more appropriate
  Bayesian cumulative normal statistical model can more accurately distinguish
  group polarization from simple consensus. Group polarization continues to 
  pose a problem for psychological theory with an ever-expanding set of potential theoretical
  explanations. A search for explanations is pointless, however, without sound
  evidence. We believe this work will provide a sturdier analytical foundation for 
  future group-polarization research.
\end{abstract}
\clearpage

@@

\begin{itemize}
  \item 
    \emph{Group polarization} is a process where the mean opinion of a 
    like-minded group becomes more extreme over time~\cite{Brown1986,Brown2000}. 
    It is therefore an important component process in
    the broader process of increasing political polarization that can hamstring
    governments~\cite{Mason2018UncivilAgreementBook,Klein2020}.
  \item
    Individuals' opinions in group polarization experiments are typically 
    measured using an ordinal scale, such
    as a Likert scale, where behavioral experiment participants' internal, theoretically
    continuous, opinions are converted to one of several opinion bins indiciating
    their degree of agreement or disagreement with some prompt statement.
  \item
    Recently, however, it has been shown that it is critical to use 
    appropriate statistical methods when using experimental designs that 
    measure continuous latent psychological variables with ordinal
    measurements---failure to do so can result in false positive support for
    the presence of experimental effects~\cite{Liddell2018}.
  \item
    In this paper, we develop a generative modeling approach for evaluating whether or
    not published detections of group polarization are plausibly false, i.e.,
    whether they are \emph{plausibly-false positives} (PFPs). 
  \item
    We found significant evidence that many published group polarization findings
    are indeed plausibly false since we found distribution parameters that
    generated simulated pre- and post-discussion data that a $t$-test found
    to be from distributions with different means, although there was no difference
    in mean, only a difference in variance. 
  \item
    Although a $t$-test may be fooled by simple consensus under ordinal measurements,
    we show that a more appropriate ordered probit model of the data yield no
    such plausibly-false positives.
  \item
    Our work shows the need for using appropriate statistical tests, as well as
    the need for a consistent experimental paradigm so that meaningful meta-analyses
    can be performed for theoretical integration and rigorous understanding of
    the group polarization effect, should it prove to occur in future group
    polarization studies that use proper statistics. In addition to avoiding
    $t$-tests, future group polarization studies can be more reliable by
    accounting for random effects arising from, e.g., socioeconomic status,
    gender, and other individual-level factors that may bias how much and in what
    direction a participant changes their opinions. Failure to do so can result 
    in artificially inflated effect sizes~\cite{Yarkoni2021}.
\end{itemize}

@@

Group polarization is often induced in behavioral experiments by first answering
an opinion prompt, grouping
relatively like-minded individuals together to discuss the prompt, then 
again polling opinions; the change in mean opinion towards a more extreme
consensus than the initial mean is the amount of group polarization that has occurred
(Figure~\ref{fig:groupPolarizationSchematic}).

While all ten studies scrutinized here use the general experimental procedure
outlined above (Figure~\ref{fig:groupPolarizationSchematic}), each study and constituent 
experimental condition have idosyncracies that makes the results themselves
more opaque, and that makes comparison of different studies' findings more
difficult. (REVIEW A FEW SIMPLE EXAMPLES). A detailed examination of each study's
idiosyncracies is given in Table~\ref{tab:conditionsComparison}.

\begin{figure}
  \caption{Group polarization is experimentally induced by first gather
  participant responses to an opinion prompt and grouping together like-minded
  participants (A). Group polarization is theorized to occur when individuals
  are put together with like-minded individuals discuss the prompt (B). Group 
  polarization is detected by again polling all participants' opinions (C) and comparing
  these to the initial opinions.}
  \label{fig:groupPolarizationSchematic}
  \centering
  \includegraphics[width=\textwidth]{Figures/GPExperimentSketch.pdf}
\end{figure}

\spacing{1.0}
\input{tables/conditions_table.tex}
\spacing{1.5}
